Qualitative Assessesment of Graph Neural Network Potentials for Phonon Properties of Shape Memory Alloys using Graph Neural Network Potentials

### 2024-01-18 src/tex/ms.tex
Updates to plot scripts

%The MACE potential pretrained \texttt{2023-12-03-mace-128-L1_epoch-199.model} checkpoint file.

@@

\begin{equation}
  \label{eq:cohesive_energy}
  E_{c} = \frac{E_{\text{total}}}{N}
\end{equation}

With $E_{\text{total}}$ the total energy of the system and $N$ the number of atoms. In fig.~\ref{fig:ecoh} the bar chart for different phases and interatomic potentials models are show. We compare these to the predicted cohesive energy for The Materials Project (MP) entry \texttt{MP-571} which corresponds to the \textit{B2} phase \cite{MP--571}. In the figure we see that the EAM and MEAM potentials show significantly lower cohesive energy compared to the GNN potentials. The stark increase in binding energy for the GNN is surprising as the EAM and MEAM potentials are known to reproduce previous experimental and DFT data \cite{Haskins2016,Ko2015}. The GNNs predict a more strongly bound and stable \textit{B2} structure but all are consistent with the entry for The MP and calculation details can be found in ref.~\cite{MP--571}. \par

\begin{figure}[ht!]
    \script{PlotCohesiveEnergy.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/NiTi_CohesiveEnergyPlot.png}
        \caption{
          Comparison of predicted \ce{NiTi} cohesive energy for respective ground-state structure and model. As a reference we use The Materials Project entry for the \ce{NiTi} \textit{B2} entry \cite{MP--571}.
        }
        \label{fig:ecoh_niti}
    \end{centering}
\end{figure}

@@

\subsection{NiTi: Phonon Instability}
\label{subsec:niphonons}

\subsubsection{B2 Phase}
\label{subsubsec:b2}

@@

\begin{figure}[!htp]
    \script{PlotPhonons.py}
    \begin{centering}
        \includegraphics[width=0.5\linewidth]{figures/NiTi_Mutter_B2_StrainsPhononBandstructures.png}
        \vspace{1mm}
        \includegraphics[width=0.5\linewidth]{figures/NiTi_Zhong_B2_StrainsPhononBandstructures.png}
        \vspace{1mm}
        \includegraphics[width=0.5\linewidth]{figures/NiTi_Ko_B2_StrainsPhononBandstructures.png}
        \caption{
           Phonon dispersion of B2 structure for the  (top) EAM Mutter, (middle) EAM Zhong, and (bottom) 2NN-MEAM Ko potentials at different isotropic strains. The optical branches have been made more transparent to emphasis the behavior in the TA$_1$ and TA$_2$ branches. The color indicates direction from compressive to tensile strain. Note that the strain percent shown is with respect to isotropic changes (i.e. volume). 
        }
        \label{fig:mutter_zhong_phonon_b2}
    \end{centering}
\end{figure}

%ADD Plot of zero strain EAM and first-pricinpals results

The B2 phase of NiTi exhibits notable phonon instabilities, particularly at the \textit{M} point in the Brillouin zone. These instabilities are characterized by imaginary modes, represented as negative phonon frequencies, indicating a dynamical instability in this crystal structure. The \textit{M} point, located at the edge of the Brillouin zone, corresponds to a wave vector of $\left[\frac{1}{2}, \frac{1}{2}, 0\right]$ in units of the reciprocal lattice vectors. The phonon dispersion curves for NiTi in the B2 phase reveal a softening of the acoustic modes at the \textit{M} point, suggesting a propensity for a structural phase transition. \par

@@

where $V$ is the volume, usually taken at equilibrium, and $\omega$ is the phonon frequency. For the B2 phase of NiTi, the negative phonon frequencies at the M point indicate an instability that can be modulated by applying strain. The strain perturbs the lattice volume, thereby altering the phonon frequencies. The effect of this strain can be quantitatively assessed using the mode Gruneisen parameter, which describes how the phonon frequencies change with volume. A positive $\gamma$ suggests that the phonon frequency increases as the volume decreases under strain, potentially stabilizing the lattice. In the case of NiTi, applying compressive strain may increase the frequencies of the soft modes at the M point, thus stabilizing the B2 phase and delaying or suppressing the martensitic transformation. This interplay between strain and phonon behavior, quantified by $\gamma$, is crucial in tailoring the mechanical properties of shape memory and superelastic materials.

\variable{output/Table_NiTi_M_ModeGruneisen.tex}

>>>>

% The MACE potential pretrained
% \texttt{2023-12-03-mace-128-L1_epoch-199.model} checkpoint file.

@@

\begin{equation}
  \label{eq:cohesive_energy}
  E_{c} = \frac{E_{\text{bulk}}-\sum_i^N n_i E^_{\text{atom_i}}}{N}
\end{equation}

With $E_{\text{bulk}}$ the total energy of the system,
$\sum_i^N n_i E_{\text{atom_i}}$ is the sum over isolated atom
energies for $n_i$ species, and $N$ the number of atoms in bulk
reference. Classical interatomic potentials are offset with respect to
the isolated atom energies, so these terms are zero and neglected,
giving $E_{c} = \frac{E_{\text{bulk}}}{N}$. In fig.~\ref{fig:ecoh} the
bar chart for different phases and interatomic potentials models are
show. We compare these to dervided experimental values for
\textit{B2}\cite{Vandermause2024} and the value from The Materials
Project (MP) entry \texttt{MP-571} which corresponds to the
\textit{B2} phase \cite{MP--571}. In the figure we see that the EAM
and MEAM potentials show significantly lower cohesive energy compared
to the GNN potentials and are in better agreement with the
experimental value. Previous studies using the EAM and MEAM potentials
have been shown to reproduce previous experimental and DFT data
\cite{Haskins2016,Ko2015} such as cohesive energy. The GNNs predict a
more strongly bound and stable \textit{B2} structure but all are
consistent with the entry for The MP and calculation details can be
found in ref.~\cite{MP--571}. This consistency is resonable
considering all pretrained GNN models used in this work have been
developed with MP data. \par

\begin{figure}[ht!]
  \script{PlotCohesiveEnergy.py}
  \begin{centering}
    \includegraphics[width=\linewidth]{figures/NiTi_CohesiveEnergyPlot.png}
    \caption{ Comparison of predicted \ce{NiTi} cohesive energy for
      respective ground-state structure and model. The experimental
      data point has been extracted from ref.~\cite{Vandermause2024}
      and the Materials Project entry for the \ce{NiTi} \textit{B2}
      structure \cite{MP--571} has also been included.  }
    \label{fig:ecoh_niti}
  \end{centering}
\end{figure}

For the \ce{PtTi} system we only consider the \textit{B2} and
\textit{B19'} structures. The MEAM potential shows a lower cohesive
energy than all three GNN. Interestingly the difference between the
MEAM and GNN is roughly the same order, $\approx 2.1$ eV, which maybe
appears to be related to the MP dataset. To our knowledge most the
pretrained GNN models are using the MP dataset which defaults to DFT
calculation results that employ the meta-GGA R2SCAN functional. This
could be the reason for the difference as most EAM and MEAM potentials
are either parameterized from experimental or DFT data using the PBE
functional\cite{Mutter2010,Ko2015}. \par

@@

\subsection{NiTi: Phonon Instability}
\label{subsec:nitiphonons}

@@

\subsubsection{Mode Grun\"{e}isen Parameter}

The B2 phase of NiTi exhibits notable phonon instabilities,
particularly at the \textit{M} point in the Brillouin zone. These
instabilities are characterized by imaginary modes, represented as
negative phonon frequencies, indicating a dynamical instability in
this crystal structure. The \textit{M} point, located at the edge of
the Brillouin zone, corresponds to a wave vector of
$\left[\frac{1}{2}, \frac{1}{2}, 0\right]$ in units of the reciprocal
lattice vectors. The phonon dispersion curves for \ce{NiTi} in the B2
phase reveal a softening of the acoustic modes at the \textit{M}
point, suggesting a propensity for a structural phase transition. \par

@@

where $V$ is the volume, usually taken at equilibrium, and $\omega$ is
the phonon frequency. For the \textit{B2} phase of \ce{NiTi}, the
negative phonon frequencies at the \textit{M} point indicate an
instability that can be modulated by applying strain. The strain
perturbs the lattice volume, thereby altering the phonon
frequencies. The effect of this strain can be quantitatively assessed
using the mode Gr\"{u}neisen parameter, which describes how the phonon
frequencies change with volume. A positive $\gamma$ suggests that the
phonon frequency increases as the volume decreases under strain,
potentially stabilizing the lattice. In the case of NiTi, applying
compressive strain may increase the frequencies of the soft modes at
the \textit{M} point, thus stabilizing the \textit{B2} phase and
delaying or suppressing the martensitic transformation. This interplay
between strain and phonon behavior, quantified by $\gamma$, is crucial
in tailoring the mechanical properties of shape memory and
superelastic materials.

In fig.~\ref{fig:modegruneisen} the Gr\"{u}neisen parameter for
\textit{M}-mode in \textit{B2} is shown for different interatomic
potentials for \ce{NiTi}. The instable modes (i.e., negative
eigenfrequencies) for the MEAM, M3GNet, and MACE potentials have
Gr\"{u}neisen parameter which are negative. The negative values
suggest that compressive isotropic strains will provide stabalizing
force against displacive phase transformation. At higher frequencies
the Gr\"{u}neisen parameter is positive, indicating this modes
experience a decrease in frequency with respect to volume
increase. The optical modes are less sensitive as indicated by smaller
values. \par

\begin{figure}[!htp]
  \script{GruneisenParameters.py} \centering
  \includegraphics[width=\columnwidth]{figures/Plot_NiTi_M_ModeGruneisen.png}
  \label{fig:modegruneisen}
  \caption{The sign-log of the \textit{M}-mode Gr\"{u}neisen
    parameter, $\gamma$, as a function of eigenfrequency for different
    interatomic potentials.}
\end{figure}

The Mutter EAM potential shows negative Gr\"{u}''neisen parameters for
all eigenfrequencies indicating compressive forces increase
frequencies, however, none of the \textit{M}-point modes have negative
eigenfrequencies. \par

In contrast the the MEAM, M3GNet, and MACE potentials, the Zhong EAM
and CHGNet potentials show positive Gr\"{u}niesen parameters for
negatives eignefrequencies, suggesting that isotropic tensile strain
acts to stabalize the mode. \par

To better coberate the results in fig.~\ref{fig:modegruneisen}, the
phonon bandstructures as a result of isoptroic strain applied in
compression and tension are show in fig

\begin{figure}[!htp]
  \script{PlotPhonons.py}
  \begin{centering}
    \includegraphics[width=0.5\linewidth]{figures/NiTi_Mutter_B2_StrainsPhononBandstructures.png}
    \vspace{1mm}
    \includegraphics[width=0.5\linewidth]{figures/NiTi_Zhong_B2_StrainsPhononBandstructures.png}
    \vspace{1mm}
    \includegraphics[width=0.5\linewidth]{figures/NiTi_Ko_B2_StrainsPhononBandstructures.png}
    \caption{ Phonon dispersion of B2 structure for the (top) EAM
      Mutter, (middle) EAM Zhong, and (bottom) 2NN-MEAM Ko potentials
      at different isotropic strains. The optical branches have been
      made more transparent to emphasis the behavior in the TA$_1$ and
      TA$_2$ branches. The color indicates direction from compressive
      to tensile strain. Note that the strain percent shown is with
      respect to isotropic changes (i.e. volume).  }
    \label{fig:mutter_zhong_ko_phonon_b2}
  \end{centering}
\end{figure}

% ADD Plot of zero strain EAM and first-pricinpals results

% \variable{output/Table_NiTi_M_ModeGruneisen.tex}
